---
title: "Hypothesis Testing in (Dis)Similarity of Ecological Communities"
author: "Oleksii Dubovyk"
date: "5/17/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hypothesis Testing in (Dis)Similarity of Ecological Communities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
_oadubovyk@gmail.com_

## The Basics of Similarity Indices

There are many different ways to describe an ecological community: by a mere number of species or using some "diversity" index (Simpson, Shannon; Hurblert's PIE; Hill's numbers) that not always has something to do with ecology (see Hurlbert 1971^[Hurlbert, S.H. (1971) The nonconcept of species diversity: A critique and alternative parameters. _Ecology 52_(4), 577-586. https://doi.org/10.2307/1934145], Gotelli & Ellison 2018^[Gotelli, N.J., & Ellison, A.M. (2018). _A Primer of Ecological Statistics_ (2nd ed.). New York: Oxford University Press.]). Yet, the most informative way to describe a community is the raw community structure data, i.e., the metrics of abundance (density, number of individuals etc.) referred to particular taxa. 

It is possible to compare the taxonomic composition or structure of ecological communities using a variety of indices (Wolda 1981)^[Wolda, H. (1981) Similarity indices, sample size and diversity. _Oecologia 50_, 296-302. https://doi.org/10.1007/BF00344966)]. This way is more preferable because comparing a mere community parameter (some diversity index) is meaningless: two communities composed of totally different taxa can have identical diversity metric.

There are two ways to compare the communities that differ in the way the data is being held. The first group of indices treat the community structure as presence-absence data. The most popular indices are Jaccard and Sorensen:

$S_{Jaccard} = \frac{s_{A \cup B}}{s_A + s_B - s_{A \cup B}}$,

where

- $s_{A \cup B}$ - number of species common to both communities,
- $s_A$ - number of species unique to the first community ($A$), and
- $s_B$ - number of species unique to the second community ($B$);

$S_{Sorensen} = \frac{2 \cdot s_{A \cup B}}{s_A + s_B}$

where the notation is the same.

The second way treats the community structure data directly. For example, one could imagine a community as a point in multidimensional space of taxa abundances, thus, the squared Euclidian distance can be measured between the communities (points):

$1 - D^2 = 1 - \sum_{i=1}^{S} n_{A,i}^2 - n_{B,i}^2$

where $n_{j, i}$ - abundance metric value of taxa $i$ in community $j$.

Another popular metric is Bray-Curtis index:

$1 - BC = 1 - \frac{\sum_{i=1}^{S} |n_{A, i} - n_{B, i}|}{\sum_{i=1}^{S} n_{A, i} + n_{B, i}}$

where the notation is the same.

It is easy to write the functions for every possible similarity index in R. I have done this several years ago and united them within a package [`ancends`](https://github.com/OleksiiDubovyk/ancends/blob/master/ancends_0.1.1.tar.gz?raw=true) (honestly, I do not remember exactly what did I want to say with this name). To use the code here you would need to download this package as an archive and to install it (RStudio > Tools > Install Packages > Install from: Package Archive File (.tar.gz)).

```{r}
library(ancends)
```

As an example dataset, you will need to use vectors of taxa abundances. I suggest using my raw data^[Dubovyk, O., Kuzyo, H., & Bokotey, A. (2020). Density variation in "rare" breeding birds in native forests and urban parks. _Geo&Bio 19_, 20-31. http://doi.org/10.15407/gb1904] on abundances of 67 bird species at 59 500-m transects among 16 forested areas in and around Lviv, Ukraine:

```{r}
lbt <- read.csv("https://github.com/OleksiiDubovyk/test/files/6498008/lbt.csv")
rownames(lbt) <- lbt$Transect
lbt <- lbt[,-1]
head(lbt)
```

The syntax for comparison of two vector is simple. For example, to calculate Jaccard index for communities of transects BC1 and BC2 (identical first two letters imply that the transects were located close to each other at the same area) we type:

```{r}
S_j <- jaccard(data.frame(lbt$BP1, lbt$BP2))
S_bc <- braycurtis(data.frame(lbt$BP1, lbt$BP2))
similarities <- c(S_j, S_bc)
names(similarities) <- c("Jaccard", "Bray-Curtis")
similarities
```

It means that we have 62.9% of taxonomic overlap between two communities and their structures, according to Bray-Curtis index, are 79.6% similar.

Of course, we can calculate these values manually to validate the results. Firstly, take a look at these two vectors:

```{r}
lbt[,1:2]
```

Next, we should calculate the number of shared species $s_{A \cup B}$ and total numbers of species in both communities ($s_A$ and $s_B$):

```{r}
s_A <- sum(lbt$BP1 > 0)
s_B <- sum(lbt$BP2 > 0)
s_AB <- sum(sapply(1:nrow(lbt), function(i){
  lbt$BP1[i] > 0 & lbt$BP2[i] > 0 
}))
s <- c(s_AB, s_A, s_B)
names(s) <- c("s_AB", "s_A", "s_B")
s
```

We see that $S_{Jaccard} = \frac{s_{A \cup B}}{s_A + s_B - s_{A \cup B}} = \frac{17}{21+23-17}$ indeed equals $0.6296296$.

Recreating Bray-Curtis index calulation is a little bit more complicated:

```{r}
cbind(lbt$BP1, lbt$BP2, abs(lbt$BP1 - lbt$BP2), lbt$BP1 + lbt$BP2)

#first sum
sum(abs(lbt$BP1 - lbt$BP2))
#second sum
sum(lbt$BP1 + lbt$BP2)
```
Again, $1 - BC = 1 - \frac{\sum_{i=1}^{S} |n_{A, i} - n_{B, i}|}{\sum_{i=1}^{S} n_{A, i} + n_{B, i}} = 1 - \frac{19}{93} = 1 - 0.2043011 =0.7956989$ which is exactly what we needed to show. Thus we believe that we can trust the function defined in `ancends`, at least `jaccard` and `braycurtis`.

## Hypothesis Testing

But what is the meaning of those indices? When can we say that two communities are different? Below 50%? But this cut-off level would not make sense due to the fact that different indices yield different values with identical data.

We can approach this question from the perspective of classical frequentist-based hypothesis testing: p-value corresponding to the probability of rejecting the null hypothesis when it is true would be useful here. But firstly we should define the statistical hypotheses.

Let $a$ and $b$ be two sets of labeled objects (individuals that belong to some taxa $i$). These two sets are sampled from the big sets $A$ and $B$. Since we are interested whether $a$ and $b$ are significantly different, we may define the hypotheses that

* The Null hypothesis $H_0$:
  * the samples were sampled from the same sets, i.e., $a \subset A = B \supset b$
* The Alternative hypothesis $H_A$:
  * the samples were sampled from the different sets, i.e., $a \subset A \neq B \supset b$.

Thus, we would need to yield a distribution of a similarity index under the null hypothesis (two samples were taken from a shared “metacommunity” pool).

For example, for two samples

```{r}
df <- lbt[,1:2]
```

the pool of labeled individuals would be

```{r}
pool <- unlist(sapply(1:nrow(df), function(i){
  sapply(1:2, function(j){
    rep(rownames(df[i,]), df[i, j])
  })
}))
sort(pool)
```

so that there is one individual of the Long-Tailed Tit Aegithalos caudatus, Short-Toed Treecreeper Certhia brachydactyla, Eurasian Treecreeper Certhia familiaris two Hawfinches Coccothraustes coccothraustes and so on.

Under the null hypothesis, the samples were collected from this pool so that the first community would be sampled to collect the number of individuals that have been collected in the original vector:

```{r}
n1 <- sum(df[1])
n1
```
```{r}
p1 <- sample(x = pool, size = n1)
p1
```

```{r}
ps1 <- sapply(rownames(df), function(i){
  sum(sapply(p1, function(j){
    j == i
  }))
})
ps1
```

so that we have a sample expected to be collected under the null hypothesis for the first vector. This can be repeated for the second vector:

```{r}
n2 <- sum(df[2])
p2 <- sample(x = pool, size = n2)
ps2 <- sapply(rownames(df), function(i){
  sum(sapply(p2, function(j){
    j == i
  }))
})
```

and we have two samples to be compared with some similarity index:

```{r}
pseudo_df <- data.frame(ps1, ps2)
jaccard(pseudo_df)
```

After repeating this operation many times, a distribution of the index can be inferred.

```{r}
resample.communities <- function(com1, com2){
  df <- data.frame(com1 = com1, com2 = com2)
  pool <- unlist(sapply(1:nrow(df), function(i){
    sapply(1:2, function(j){
      rep(rownames(df[i,]), df[i, j])
    })
  }))
  n1 <- sum(df[1])
  n2 <- sum(df[2])
  perm1 <- sample(x = pool, size = n1)
  perm2 <- sample(x = pool, size = n2)
  pseudo1 <- sapply(rownames(df), function(i){
    sum(sapply(perm1, function(j){
      j == i
    }))
  })
  pseudo2 <- sapply(rownames(df), function(i){
    sum(sapply(perm2, function(j){
      j == i
    }))
  })
  jaccard(data.frame(pseudo1, pseudo2))
}
```

```{r}
distr <- numeric(1000)
for (i in 1:1000){
  distr[i] <- resample.communities(lbt[,1], lbt[,2])
}
hist(distr, freq = F)
lines(density(distr))
```

It is known that the observed Jaccard index equals 0.6296296. Since under the null hypothesis the two communities are similar, i.e., a similarity metric should $S \rightarrow 1$, the extreme values of the metric are lower-tailed. Thus, the p-value in this case is probability of observation of the observed metric value or lower.

```{r}
sum(distr <= 0.6296296)/1000
```

```{r}
hist(distr, freq = F)
lines(density(distr))
abline(v = 0.6296296, col = "red", lwd = 3)
```

In that case, there is not enough evidence to decide that these two communities were sampled from different metacommunities. It makes sense, because the data were collected from the transects few hundred meters apart in the same forest.

We can define a function to use different indices. It takes two vectors corresponding to community structures (note that two vectors have to be of the same length and their indices should correspond to the same taxa) as arguments `com1` and `com2`, `nperm = 1000` defines the number of permutations, `distance` utilizes a function of some similarity index (the functions from `ancends` are supposed to be used). The result can be yielded as a vector of similarity index value and p-value if `show.hist = FALSE` or as histogram if `show.hist = TRUE`. It takes approximately 40 seconds for 1000 permutations.

```{r}
compare.communities <- function(com1, com2, nperm = 1000, distance = ancends::jaccard, show.hist = F){
  fnct <- distance
  resample.communities <- function(com1 = com1, com2 = com2){
    df <- data.frame(com1 = com1, com2 = com2)
    pool <- unlist(sapply(1:nrow(df), function(i){
      sapply(1:2, function(j){
        rep(rownames(df[i,]), df[i, j])
      })
    }))
    n1 <- sum(df[1])
    n2 <- sum(df[2])
    perm1 <- sample(x = pool, size = n1)
    perm2 <- sample(x = pool, size = n2)
    pseudo1 <- sapply(rownames(df), function(i){
      sum(sapply(perm1, function(j){
        j == i
      }))
    })
    pseudo2 <- sapply(rownames(df), function(i){
      sum(sapply(perm2, function(j){
        j == i
      }))
    })
    df <- data.frame(pseudo1, pseudo2)
    fnct(df)
  }
  distr <- numeric(nperm)
  for (i in 1:nperm){
    distr[i] <- resample.communities(lbt[,1], lbt[,2])
  }
  metric <- fnct(data.frame(com1 = com1, com2 = com2))
  rslt <- c(round(metric, 3), ifelse(sum(distr <= metric)/nperm == 0, 
                                     paste("<", 1/nperm), 
                                     sum(distr <= metric)/nperm))
  names(rslt) <- c("Metric", "p-value")
  if (show.hist){
    hist(distr, main = paste("Metric = ", rslt[1], "; p-value = ", rslt[2], sep = ""), 
         freq = F, xlab = "Metric", xlim = c(0, 1))
    lines(density(distr), col = "blue")
    abline(v = rslt[1], col = "red", lwd = 3)
  }else{
    rslt
  }
}
```

The example of use:

```{r}
compare.communities(lbt[1], lbt[5], nperm = 1000, distance = ancends::braycurtis, show.hist = T)
compare.communities(lbt[1], lbt[10], nperm = 1000, distance = ancends::jaccard, show.hist = T)
compare.communities(lbt[1], lbt[15], nperm = 1000, distance = ancends::jaccard, show.hist = F)
compare.communities(lbt[1], lbt[20], nperm = 1000, distance = ancends::braycurtis, show.hist = F)
```

